---
title: "Course Project"
author: "Miguel González Lugo"
date: "7/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Course Project 2
This r markdown performs an exploratory analysis on data from a personal monitoring device. This device tracked the steps of an anonimous person in intervals of 5 minutes. 

## Reading and manipulationg the data

The data is read using read.csv, the "date" variable is change into a variable of class data using lubridate package
```{r, results='hide'}
library(lubridate)
```


```{r reading data}
actData <- read.csv("activity.csv")
actData$date <- ymd(actData$date)
```

## Histogram of the total number of steps taken each day

Grouping by date.
```{r histogram}
DaySteps <- with(actData, tapply(steps, date, sum, na.rm = T))
hist(DaySteps)
```

## Mean and median number of steps taken each day

```{r mean and median}
sum <- summary(DaySteps)
sum
```
The median was of `r sum[[3]]` and the mean of `sum[[4]]`


## Time series plot of the average number of steps taken

```{r}
AvgStp <- with(actData, tapply(steps, date, mean, na.rm = T))
plot(AvgStp~as.Date(names(AvgStp)), main = "Average Steps in Time", type = "l", ylab = "Avg Steps", xlab = "")
```

## The 5-minute interval that, on average, contains the maximum number of steps

```{r}
actData$interval <- as.factor(actData$interval)
AvgStpFreq <- with(actData, tapply(steps, interval, mean, na.rm = T))
max <- AvgStpFreq[which.max(AvgStpFreq)[[1]]]
max
```
On average in all days the interval `r names(max)` had a maximum number of steps of `r max[[1]]`

## Code to describe and show a strategy for imputing missing data.

The following code loops over each observation in 'steps' and, if it's NA, replaces the observation with the mean of all the observation with the same interval.
```{r missing values}
c <- colSums(is.na(actData))
for (i in 1:nrow(actData)) {
        if (is.na(actData$steps[i])) {
                int <- as.numeric(as.character(actData$interval[i]))
                actData$steps[i] <- mean(subset(actData, interval == int)$steps, na.rm = T)
        }
}
```
Intially, before imputing NA for mean values, the numer of rows with NA was of `r c[[1]]`

## Histogram of the total number of steps taken each day after missing values are imputed

```{r}
DaySteps <- with(actData, tapply(steps, date, sum))
hist(DaySteps)
sum2 <- summary(DaySteps)
```
With the summary calculate abvoce we'll calculate it's difference with the non-imputed dataset.
```{r summary comparisson}
sumdif <- c()
for (i in 1:length(sum)) {
        sumdif[i] <- sum2[i]-sum[i]
}
names(sumdif) <- names(sum)
sumdif
```
It's observed that the minimum, 1st quartile, median and mean increased after imputing the missing values. 

## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

First we  create a variable indicating weather day is a weekday or weekend. 
```{r type variable}
actData$weekday <- factor(weekdays(actData$date))
for (i in 1:nrow(actData)) {
        if (actData$weekday[i] == "Sunday") {
                actData$Type[i] <- "Weekend"
        }
        else if (actData$weekday[i] == "Saturday") {
                actData$Type[i] <- "Weekend"
        }
        else {
                actData$Type[i] <- "Weekday"
        }
}
```
Now we group the data and plot it using the new variable type as a facet. 
```{r}
library(lattice)
DatSteps <- aggregate(steps~ interval + Type, data = actData, mean)
xyplot(steps~interval | Type, data = DatSteps, type = "l", ylab = "Avrg Steps", xlab = "Interval", scales = list(x = list(cex = 0)))
```
We see that on week days there were more steps(avrg) at lower intervals than on weekend. This may signify that on weekday this person walked more on the mornings than on afternoos. On the other hand on weekends there was less activity on the morning and more in the afternoon. 
